---
title: "Endogenous RNA mass"
author: "Franco Poma"
date: "2025-02-11"
output: html_document
---

# Load libraries

```{r}
library(dplyr)
require(readr)
library(DT)
data_dir <- "./Data"
```

# Make annotation only Nash subcohort

```{r, include=FALSE}
#Load outliers
outliers_df<-read.table(paste0(data_dir, "/OutliersC2.txt"), header = T) # set of 41 outliers.
#Annotation
ann<-read_tsv(paste0(data_dir, "/Annotation_Cohort2.tsv"))
#Select only NASH cohort and baseline samples
pDat_NASH<- ann %>% filter(str_detect(Cohort, "NASH"))
pDat_NASH <- pDat_NASH %>% dplyr::filter(Visit_Type=="Baseline")

#Removing NoNAFLD samples
pDat_NASH <- pDat_NASH %>% dplyr::filter(NASH!="No NAFLD")
colnames(pDat_NASH)[1]<-"RNAID"
#Removing Outliers
pDat_NASH<-pDat_NASH %>% dplyr::filter(!RNAID%in%outliers_df$RNAID)
#New varaible NASH_risk
pDat_NASH <- pDat_NASH %>%  mutate("Risk" = as.factor(if_else(pDat_NASH$NASH=="NASH"&pDat_NASH$Fibrosis%in%c("F2","F3", "F4", "F5", "F6"), "Risk", "No_Risk")))
```

# Load counts

```{r, warning=FALSE, message=FALSE}
htseq <- read_delim(paste0(data_dir, "/htseq_cohort2.txt"))
#remove ambiguous, not unique, no feature
htseq <- htseq %>% filter(!str_detect(ensembl_gene_id, "^__"))
#Select only Nash cohort counts
htseq <- htseq  %>%
  column_to_rownames("ensembl_gene_id")

count_table_NASH <- htseq %>% dplyr::select(pDat_NASH$RNAID)

count_table_NASH <- count_table_NASH  %>%
  rownames_to_column("ensembl_gene_id")
```

# RNA concentration

### Step 1: Calculate the Diluted Concentration of the Spike-In

If the stock concentration is $C_{\text{stock}}$ (in attomol/μL) and the dilution factor is $D$, the diluted concentration is: $$
C_{\text{diluted}} = \frac{C_{\text{stock}}}{D}
$$

### Step 2: Calculate Total Moles Added to the Lysis Buffer

Given a volume $V_{\text{added}}$ (in μL) of the diluted spike-in solution: $$
n_{\text{total}} = C_{\text{diluted}} \times V_{\text{added}} \times 10^{-18} \quad \text{(moles)}
$$ *Note:* The factor $10^{-18}$ converts attomol to mol.

### Step 3: Adjust for the Volume Transferred During Purification

If the total lysis buffer volume is $V_{\text{lysis}}$ (μL) and the volume transferred is $V_{\text{transferred}}$ (μL), then the moles in the transferred volume are: $$
n_{\text{transferred}} = n_{\text{total}} \times \frac{V_{\text{transferred}}}{V_{\text{lysis}}}
$$

### Step 4: Calculate the Molecular Weight of the Spike-In RNA

For an RNA molecule of length $L$ nucleotides, its molecular weight is: $$
MW = (L \times 321.47\ \text{g/mol}) + 18.02\ \text{g/mol}
$$

### Step 5: Calculate the Mass of the Spike-In

The mass (in grams) of the spike-in in the transferred volume is: $$
\text{Mass}_{\text{spike}} = n_{\text{transferred}} \times MW
$$ To convert this mass to nanograms (ng): $$
\text{Mass}_{\text{spike}}\, (\text{ng}) = n_{\text{transferred}} \times MW \times 10^9
$$

### Step 6: Estimate the Mass of Endogenous RNA

Using the sequencing read counts for endogenous RNA ($\text{Reads}_{\text{endogenous}}$) and for the spike-in ($\text{Reads}_{\text{spike}}$), the mass of endogenous RNA is estimated by: $$
\text{Mass}_{\text{endogenous}} = \text{Mass}_{\text{spike}} \times \frac{\text{Reads}_{\text{endogenous}}}{\text{Reads}_{\text{spike}}}
$$

```{r}
#Keep only sequin spikes
Spikes_counts <- filter(count_table_NASH, str_detect(ensembl_gene_id,"R1|R2"))

## get most abundant spike
Spikes_counts$Total <- rowSums(Spikes_counts[2:175])
Top_Spike<-Spikes_counts[order(Spikes_counts$Total, decreasing=T),][1,]
rownames(Top_Spike)<-NULL
Top_Spike <- Top_Spike[,1:ncol(Top_Spike)-1]
Top_Spike <- Top_Spike %>% column_to_rownames(var = "ensembl_gene_id")
Top_Spike_name <- rownames(Top_Spike)[1]
print (paste0("Top Spike is: ", Top_Spike_name))
```

```{r}
Top_Spike <- as.data.frame(t(Top_Spike))
Top_Spike <- Top_Spike %>% rownames_to_column(var="RNAID")

# Load spike counts and merge
Spikes_C2<- read.csv(paste0(data_dir, "/SpikeRatio_C2.csv"))


Top_Spike <- left_join(Top_Spike, Spikes_C2)

#load sum of endogenous from previous analysis
Spikes_C1<- read.csv(paste0(data_dir, "/SpikeRatio_C1.csv"))
Spikes_C2<- read.csv(paste0(data_dir, "/SpikeRatio_C2.csv"))
#Top spike is R2_65
#concentration attom/ul
spike_stock<-966.797
#length (nt)
spike_len<-908
#Sequin mix vol added to lysis buffer (ul)
spike_vol<-8.75
dilution <- 1/3000
#Lysis Buffer Volume ul
LysisVol_initial<-2500
LysisVol_Final<-2000

#moles of sequin R2_65 in 8.75ul
spike_moles <- spike_stock*dilution*spike_vol*10^-18
#Moles transferred to final mix
spike_moles_final <- spike_moles *(LysisVol_Final/LysisVol_initial)

#Molecular weight of spike (g/mol) - standard formula to calculate weight of a single stranded RNA molecule
spike_mw <- (spike_len * 321.47) + 18.02
#mass (g) added to the mix
spike_mass <- spike_moles_final * spike_mw
#mass in nanograms
spike_mass_ng <- spike_mass * 10^9
#endogenous RNA mass calculation in 2000ul
Top_Spike <- Top_Spike %>% mutate(Endo_mass=spike_mass_ng*endogenous/R2_65)
#endogenous RNA concentration per ml 
Top_Spike <- left_join(Top_Spike, pDat_NASH, by="RNAID") 
Top_Spike$Plasma_input_vol_ul  <- if_else(Top_Spike$`QC_Plasma_input_volume_(µl)` == "less than 2 ml", 1990, as.numeric(Top_Spike$`QC_Plasma_input_volume_(µl)`))
Top_Spike <- Top_Spike %>% mutate(Endo_conc=Endo_mass/Plasma_input_vol_ul)
```

# Statistical Analysis


```{r}
# png("qqplot_NashRisk.png", width = 800, height = 600)
qqnorm(Top_Spike$Endo_conc)
qqline(Top_Spike$Endo_conc, col = "red")
```

```{r}
hist(Top_Spike$Endo_conc, breaks = 20, probability = TRUE,
     main = "Histogram of Endogenous Mass", xlab = "Endogenous Mass")
lines(density(Top_Spike$Endo_conc), col = "blue")

```

### Largely skewed distribution. There are 4 samples with significant larger endogenous mass. Therefore a wilcoxon Rank-Sum for two groups will be applied.

Taking into account Hospital as batch effect
```{r}
#install.packages("coin")
library(coin)

Top_Spike$Hospital <- as.factor(Top_Spike$Hospital)
coin::wilcox_test(Endo_conc ~ Risk | Hospital, data = Top_Spike,
            distribution = "exact")
```
Without hospital
```{r}
coin::wilcox_test(Endo_conc ~ Risk, data = Top_Spike)
```
No significant differences between Nash Risk groups.