---
title: "Endogenous RNA mass"
author: "Franco Poma"
date: "2025-02-11"
output: html_document
---

# Load libraries

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
require(readr)
library(DT)
library(ggplot2)
library(stringr)
require(ggbeeswarm)
color_panel<-c("#e35d6a","#428bca","darkgray","#e87810","#5bb75b", "#23496b","#ffbf00","#cc2028","#039748","pink","gray")
theme_final<- theme(axis.text.x=element_blank(), axis.text.y=element_text(size=8),axis.title.x = element_blank(), axis.title.y = element_text(size=8),plot.title = element_text(hjust = 0.5, size = 8))
data_dir <- "./Data/"
```

# Make annotation only Nash subcohort

```{r, warning=FALSE, message=FALSE}
#Load outliers
outliers_df<-read.table(paste0(data_dir, "OutliersC2.txt"), header = T) # set of 41 outliers.
#Annotation
ann<-read_csv(paste0(data_dir, "Suplementary_Table2.csv"))
#Select only NASH cohort and baseline samples
pDat_NASH<- ann %>% filter(str_detect(Cohort, "NASH"))
pDat_NASH <- pDat_NASH %>% dplyr::filter(Visit_Type=="Baseline")

#Removing NoNAFLD samples
pDat_NASH <- pDat_NASH %>% dplyr::filter(NASH!="No NAFLD")
colnames(pDat_NASH)[1]<-"RNAID"
#Removing Outliers
pDat_NASH<-pDat_NASH %>% dplyr::filter(!RNAID%in%outliers_df$RNAID)
#New varaible NASH_risk
pDat_NASH <- pDat_NASH %>%  mutate("Risk" = as.factor(if_else(pDat_NASH$NASH=="NASH"&pDat_NASH$Fibrosis%in%c("F2","F3", "F4", "F5", "F6"), "Risk", "No_Risk")))
pDat_NASH$Fibrosis <- factor(pDat_NASH$Fibrosis, levels = c("F0", "F1", "F2", "F3", "F4"))
```

# Load counts

```{r, warning=FALSE, message=FALSE}
htseq <- read_delim(paste0(data_dir, "counts/htseq_spikes_C2.txt"))
#remove ambiguous, not unique, no feature
htseq <- htseq %>% filter(!str_detect(ensembl_gene_id, "^__"))
#Select only Nash cohort counts
htseq <- htseq  %>%
  column_to_rownames("ensembl_gene_id")

count_table_NASH <- htseq %>% dplyr::select(pDat_NASH$RNAID)

count_table_NASH <- count_table_NASH  %>%
  rownames_to_column("ensembl_gene_id")
```

# RNA concentration

### Step 1: Calculate the Diluted Concentration of the Spike-In

If the stock concentration is $C_{\text{stock}}$ (in attomol/μL) and the dilution factor is $D$, the diluted concentration is: $$
C_{\text{diluted}} = \frac{C_{\text{stock}}}{D}
$$

### Step 2: Calculate Total Moles Added to the Lysis Buffer

Given a volume $V_{\text{added}}$ (in μL) of the diluted spike-in solution: $$
n_{\text{total}} = C_{\text{diluted}} \times V_{\text{added}} \times 10^{-18} \quad \text{(moles)}
$$ *Note:* The factor $10^{-18}$ converts attomol to mol.

### Step 3: Adjust for the Volume Transferred During Purification

If the total lysis buffer volume is $V_{\text{lysis}}$ (μL) and the volume transferred is $V_{\text{transferred}}$ (μL), then the moles in the transferred volume are: $$
n_{\text{transferred}} = n_{\text{total}} \times \frac{V_{\text{transferred}}}{V_{\text{lysis}}}
$$

### Step 4: Calculate the Molecular Weight of the Spike-In RNA

For an RNA molecule of length $L$ nucleotides, its molecular weight is: $$
MW = (L \times 321.47\ \text{g/mol}) + 18.02\ \text{g/mol}
$$

### Step 5: Calculate the Mass of the Spike-In

The mass (in grams) of the spike-in in the transferred volume is: $$
\text{Mass}_{\text{spike}} = n_{\text{transferred}} \times MW
$$ To convert this mass to nanograms (ng): $$
\text{Mass}_{\text{spike}}\, (\text{ng}) = n_{\text{transferred}} \times MW \times 10^9
$$

### Step 6: Estimate the Mass of Endogenous RNA

Using the sequencing read counts for endogenous RNA ($\text{Reads}_{\text{endogenous}}$) and for the spike-in ($\text{Reads}_{\text{spike}}$), the mass of endogenous RNA is estimated by: $$
\text{Mass}_{\text{endogenous}} = \text{Mass}_{\text{spike}} \times \frac{\text{Reads}_{\text{endogenous}}}{\text{Reads}_{\text{spike}}}
$$ \

### Step 7: Convert the Mass of Endogenous RNA to Concentration (ng/ml)

Using the sequencing read counts for endogenous RNA ($\text{Reads}_{\text{endogenous}}$) and for the spike-in ($\text{Reads}_{\text{spike}}$), the mass of endogenous RNA is estimated by: $$
\text{Concentration} =\frac{\text{Mass}_{\text{endogenous}}}{\text{PlasmaVol}}
$$

```{r, warning=FALSE, message=FALSE}
#Keep only sequin spikes
Spikes_counts <- filter(count_table_NASH, str_detect(ensembl_gene_id,"R1|R2"))

## get most abundant spike
Spikes_counts$Total <- rowSums(Spikes_counts[2:175])
Top_Spike<-Spikes_counts[order(Spikes_counts$Total, decreasing=T),][1,]
rownames(Top_Spike)<-NULL
Top_Spike <- Top_Spike[,1:ncol(Top_Spike)-1]
Top_Spike <- Top_Spike %>% column_to_rownames(var = "ensembl_gene_id")
Top_Spike_name <- rownames(Top_Spike)[1]
print (paste0("Top Spike is: ", Top_Spike_name))
```

```{r, warning=FALSE, message=FALSE}
Top_Spike <- as.data.frame(t(Top_Spike))
Top_Spike <- Top_Spike %>% rownames_to_column(var="RNAID")

# Load spike counts and merge
Spikes_C2<- read.csv(paste0(data_dir, "SpikeRatio_C2.csv"))


Top_Spike <- left_join(Top_Spike, Spikes_C2)
#Top spike is R2_65
#concentration attom/ul got from provider
spike_stock<-966.797
#length (nt)
spike_len<-908
#Sequin mix vol added to lysis buffer (ul)
spike_vol<-8.75
dilution <- 1/3000
#Lysis Buffer Volume ul
LysisVol_initial<-2500
LysisVol_Final<-2000

#moles of sequin R2_65 in 8.75ul
spike_moles <- spike_stock*dilution*spike_vol*10^-18
#Moles transferred to final mix
spike_moles_final <- spike_moles *(LysisVol_Final/LysisVol_initial)

#Molecular weight of spike (g/mol) - standard formula to calculate weight of a single stranded RNA molecule
spike_mw <- (spike_len * 321.47) + 18.02
#Spike mass (g) added to the mix
spike_mass <- spike_moles_final * spike_mw
#Spike mass in nanograms
spike_mass_ng <- spike_mass * 10^9
#endogenous RNA mass(ng) in 2000ul
Top_Spike <- Top_Spike %>% mutate(Endo_mass=spike_mass_ng*endogenous/R2_65)
#endogenous RNA concentration per ul 
Top_Spike <- left_join(Top_Spike, pDat_NASH, by="RNAID") 
Top_Spike$Plasma_input_vol_ul  <- if_else(Top_Spike$`QC_Plasma_input_volume_(µl)` == "less than 2 ml", 1990, as.numeric(Top_Spike$`QC_Plasma_input_volume_(µl)`))
Top_Spike <- Top_Spike %>% mutate(Endo_conc=Endo_mass/Plasma_input_vol_ul)
datatable(Top_Spike)
```

# Statistical Analysis

## Distribution of endogenous RNA concentration

```{r, warning=FALSE, message=FALSE}
# png("qqplot_NashRisk.png", width = 800, height = 600)
qqnorm(Top_Spike$Endo_conc)
qqline(Top_Spike$Endo_conc, col = "red")
```

```{r, warning=FALSE, message=FALSE}
hist(Top_Spike$Endo_conc, breaks = 20, probability = TRUE,
     main = "Histogram of Endogenous Mass", xlab = "Endogenous Mass")
lines(density(Top_Spike$Endo_conc), col = "blue")

```

**Largely skewed distribution. There are 4 samples with significant larger endogenous RNA concentration. Therefore a wilcoxon Rank-Sum for two groups will be applied.**

#### Nash Risk

Taking into account Hospital as batch effect

```{r, warning=FALSE, message=FALSE}
#install.packages("coin")
library(coin)

Top_Spike$Hospital <- as.factor(Top_Spike$Hospital)
coin::wilcox_test(Endo_conc ~ Risk | Hospital, data = Top_Spike,
            distribution = "exact")
```

Without hospital

```{r}
coin::wilcox_test(Endo_conc ~ Risk, data = Top_Spike)
```

No significant differences between Nash Risk groups.

#### Fibrosis

Taking into account Hospital as batch effect

```{r, warning=FALSE, message=FALSE}
data <- Top_Spike %>% dplyr::filter(Fibrosis%in% c("F0", "F4"))%>%
  droplevels()
coin::wilcox_test(Endo_conc ~ Fibrosis | Hospital, data = data,
            distribution = "exact")
```

Without hospital

```{r, warning=FALSE, message=FALSE}
coin::wilcox_test(Endo_conc ~ Fibrosis, data = data)
```

Both tests show a significant difference in the endogenous RNA concentration between Fibrosis F0 and F4 groups.

```{r, warning=FALSE, message=FALSE}
res_df <- data %>% dplyr::group_by(Fibrosis) %>% summarise(mean= sprintf("%.2e", mean(Endo_conc)), sd= sprintf("%.2e", sd(Endo_conc)))

datatable(res_df)
```

**Visualizing groups**

```{r, warning=FALSE, message=FALSE}


fib_plt1 <- data %>%  ggplot(aes(x=Fibrosis, y=Endo_conc, color=Fibrosis)) +geom_boxplot() + 
  geom_beeswarm(alpha=0.3, size = 0.05, dodge.width = 0.8, cex = 0.3) + scale_color_manual(values = color_panel)+
  geom_beeswarm(alpha=0.3, size = 0.5, dodge.width = 0.8, cex = 0.3) +
    theme_bw() + theme_final + ylab("EndogenousRNA(ng/ml)")

fib_plt1

```
