---
title: "Endogenous RNA mass"
author: "Franco Poma"
date: "2025-02-11"
output: html_document
---

# Load libraries

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
require(readr)
library(DT)
library(ggplot2)
library(stringr)
require(ggbeeswarm)
color_panel<-c("#e35d6a","#428bca","darkgray","#e87810","#5bb75b", "#23496b","#ffbf00","#cc2028","#039748","pink","gray")
theme_final<- theme(axis.text.x=element_blank(), axis.text.y=element_text(size=8),axis.title.x = element_blank(), axis.title.y = element_text(size=8),plot.title = element_text(hjust = 0.5, size = 8))
data_dir <- "./Data/"
```

# Make annotation only Nash subcohort

```{r, warning=FALSE, message=FALSE}
#annotation
ann_retro<-read_csv(paste0(data_dir,"annotation_Cohort1.csv"))
ann_retro$`QC_Plasma_input_volume_(µl)` <- 200


```

# Load counts

```{r}
#getting gene annotation
ensembl <- useEnsembl("ensembl",dataset="hsapiens_gene_ensembl", version = 109)
genes_ens <- getBM(attributes=c('ensembl_gene_id','gene_biotype','chromosome_name'),mart=ensembl)
htseq <- read_delim(paste0(data_dir, "/htseq_cohort1.txt"))
#Select only Nash cohort counts
htseq <- htseq  %>%
  column_to_rownames("ensembl_gene_id")

count_table_NASH <- htseq %>% dplyr::select(ann_retro$SMAPLE_RNA)

count_table_NASH <- count_table_NASH  %>%
  rownames_to_column("ensembl_gene_id")
```

# RNA concentration

### Step 1: Calculate the Diluted Concentration of the Spike-In

If the stock concentration is $C_{\text{stock}}$ (in attomol/μL) and the dilution factor is $D$, the diluted concentration is: $$
C_{\text{diluted}} = \frac{C_{\text{stock}}}{D}
$$

### Step 2: Calculate Total Moles Added to the Lysis Buffer

Given a volume $V_{\text{added}}$ (in μL) of the diluted spike-in solution: $$
n_{\text{total}} = C_{\text{diluted}} \times V_{\text{added}} \times 10^{-18} \quad \text{(moles)}
$$ *Note:* The factor $10^{-18}$ converts attomol to mol.

### Step 3: Adjust for the Volume Transferred During Purification

If the total lysis buffer volume is $V_{\text{lysis}}$ (μL) and the volume transferred is $V_{\text{transferred}}$ (μL), then the moles in the transferred volume are: $$
n_{\text{transferred}} = n_{\text{total}} \times \frac{V_{\text{transferred}}}{V_{\text{lysis}}}
$$

### Step 4: Calculate the Molecular Weight of the Spike-In RNA

For an RNA molecule of length $L$ nucleotides, its molecular weight is: $$
MW = (L \times 321.47\ \text{g/mol}) + 18.02\ \text{g/mol}
$$

### Step 5: Calculate the Mass of the Spike-In

The mass (in grams) of the spike-in in the transferred volume is: $$
\text{Mass}_{\text{spike}} = n_{\text{transferred}} \times MW
$$ To convert this mass to nanograms (ng): $$
\text{Mass}_{\text{spike}}\, (\text{ng}) = n_{\text{transferred}} \times MW \times 10^9
$$

### Step 6: Estimate the Mass of Endogenous RNA

Using the sequencing read counts for endogenous RNA ($\text{Reads}_{\text{endogenous}}$) and for the spike-in ($\text{Reads}_{\text{spike}}$), the mass of endogenous RNA is estimated by: $$
\text{Mass}_{\text{endogenous}} = \text{Mass}_{\text{spike}} \times \frac{\text{Reads}_{\text{endogenous}}}{\text{Reads}_{\text{spike}}}
$$\

### Step 7: Convert the Mass of Endogenous RNA to Concentration (ng/ml)

Using the sequencing read counts for endogenous RNA ($\text{Reads}_{\text{endogenous}}$) and for the spike-in ($\text{Reads}_{\text{spike}}$), the mass of endogenous RNA is estimated by: $$
\text{Concentration} =\frac{\text{Mass}_{\text{endogenous}}}{\text{PlasmaVol}}
$$

```{r, warning=FALSE, message=FALSE}
#Keep only sequin spikes
Spikes_counts <- filter(count_table_NASH, str_detect(ensembl_gene_id,"R1|R2"))
## get most abundant spike
Spikes_counts$Total <- rowSums(Spikes_counts[2:ncol(Spikes_counts)])
Top_Spike<-Spikes_counts[order(Spikes_counts$Total, decreasing=T),][1,]
rownames(Top_Spike)<-NULL
Top_Spike <- Top_Spike[,1:ncol(Top_Spike)-1]
Top_Spike <- Top_Spike %>% column_to_rownames(var = "ensembl_gene_id")
Top_Spike_name <- rownames(Top_Spike)[1]
print (paste0("Top Spike is: ", Top_Spike_name))
```

```{r, warning=FALSE, message=FALSE}
Top_Spike <- as.data.frame(t(Top_Spike))
Top_Spike <- Top_Spike %>% rownames_to_column(var="RNAID")

# Load spike counts and merge
Spikes_C1<- read.csv(paste0(data_dir, "/SpikeRatio_C1.csv"))
Top_Spike <- left_join(Top_Spike, Spikes_C1)
Top_Spike <- Top_Spike %>% dplyr::filter(R2_65!=0)


#Top spike is R2_65
#concentration attom/ul got from provider
spike_stock<-966.797
#length (nt)
spike_len<-908
#Sequin mix vol added to lysis buffer (ul)
spike_vol<-2
dilution <- 1/3000

#moles of sequin R2_65 in 2ul
spike_moles_final <- spike_stock*dilution*spike_vol*10^-18
#Molecular weight of spike (g/mol) - standard formula to calculate weight of a single stranded RNA molecule
spike_mw <- (spike_len * 321.47) + 18.02
#Spike mass (g) added to the mix
spike_mass <- spike_moles_final * spike_mw
#Spike mass in nanograms
spike_mass_ng <- spike_mass * 10^9
#endogenous RNA mass(ng) in 2000ul
Top_Spike <- Top_Spike %>% mutate(Endo_mass=spike_mass_ng*endogenous/R2_65)

#endogenous RNA concentration per ul 
Top_Spike <- left_join(Top_Spike, ann_retro, by=c("RNAID"="SMAPLE_RNA")) 
Top_Spike <- Top_Spike %>% mutate(Endo_conc=Endo_mass/`QC_Plasma_input_volume_(µl)`)

datatable(Top_Spike)
```

# Statistical Analysis

## Distribution of endogenous RNA concentration

```{r, warning=FALSE, message=FALSE}
# png("qqplot_NashRisk.png", width = 800, height = 600)
qqnorm(Top_Spike$Endo_conc)
qqline(Top_Spike$Endo_conc, col = "red")
```

```{r, warning=FALSE, message=FALSE}
hist(Top_Spike$Endo_conc, breaks = 20, probability = TRUE,
     main = "Histogram of Endogenous Mass", xlab = "Endogenous Mass")
lines(density(Top_Spike$Endo_conc), col = "blue")

```

**Largely skewed distribution. There are 4 samples with significant larger endogenous RNA concentration. Therefore a wilcoxon Rank-Sum for two groups will be applied.**

#### Cirrhosis

```{r, warning=FALSE, message=FALSE}
#install.packages("coin")
library(coin)

Top_Spike$Cirrhosis <- as.factor(Top_Spike$Cirrhosis)
coin::wilcox_test(Endo_conc ~ Cirrhosis, data = Top_Spike)
```

Significant differences between Cirrhosis groups




#### Nafld

```{r, warning=FALSE, message=FALSE}
Top_Spike$NAFLD_Diagnosis_AMC <- as.factor(Top_Spike$NAFLD_Diagnosis_AMC)
coin::wilcox_test(Endo_conc ~ NAFLD_Diagnosis_AMC, data = Top_Spike)
```

```{r, warning=FALSE, message=FALSE}
res_df <- Top_Spike %>% dplyr::group_by(Cirrhosis) %>% summarise(mean= sprintf("%.2e", mean(Endo_conc)), sd= sprintf("%.2e", sd(Endo_conc)))

datatable(res_df)
```

**Visualizing groups**

```{r, warning=FALSE, message=FALSE}


fib_plt1 <- Top_Spike[!is.na(Top_Spike$Cirrhosis),] %>%  ggplot(aes(x=Cirrhosis, y=log(Endo_conc), color=Cirrhosis)) +geom_boxplot() + 
  geom_beeswarm(alpha=0.3, size = 0.05, dodge.width = 0.8, cex = 0.3) + scale_color_manual(values = color_panel)+
  geom_beeswarm(alpha=0.3, size = 0.5, dodge.width = 0.8, cex = 0.3) +
    theme_bw() + theme_final + ylab("log EndogenousRNA(ng/ul)")

fib_plt1
ggsave("./figures/endoConc_Cirrhosis_C1.png", fib_plt1 ,dpi=300)
```
```{r}
fib_plt2 <- Top_Spike[!is.na(Top_Spike$NAFLD_Diagnosis_AMC),] %>%  ggplot(aes(x=NAFLD_Diagnosis_AMC, y=log(Endo_conc), color=NAFLD_Diagnosis_AMC)) +geom_boxplot() + 
  geom_beeswarm(alpha=0.3, size = 0.05, dodge.width = 0.8, cex = 0.3) + scale_color_manual(values = color_panel)+
  geom_beeswarm(alpha=0.3, size = 0.5, dodge.width = 0.8, cex = 0.3) +
    theme_bw() + theme_final + ylab("log EndogenousRNA(ng/ul)")

fib_plt2
ggsave("./figures/endoConc_NAFLD_C1.png", fib_plt2 ,dpi=300)
```


## ROC-AUC

```{r}
# Install and load the pROC package if you haven't already
if (!require("pROC")) {
  install.packages("pROC")
  library(pROC)
}
```


```{r}

data <- Top_Spike[!is.na(Top_Spike$Cirrhosis),]
# Ensure the cirrhosis status is numeric (if it's a factor, convert it)
data$Cirrhosis <- as.numeric(as.character(data$Cirrhosis))

# Generate the ROC curve using RNA_concentration as the predictor
roc_obj <- roc(response = data$Cirrhosis, predictor = data$Endo_conc)

png("ROC_cirrhosis_C1.png", width = 600, height = 400)
# Plot the ROC curve
plot(roc_obj, col = "#428bca", lwd = 2,
     main = "ROC Curve: RNA Concentration for Cirrhosis Prediction")
# Add a diagonal line representing random classification
abline(a = 0, b = 1, lty = 2, col = "gray")

# Calculate and print the AUC value
auc_value <- auc(roc_obj)
print(paste("The AUC is:", round(auc_value, 3)))
dev.off()
```
```{r}
# Generate the ROC curve using RNA_concentration as the predictor
roc_obj <- roc(response = data$NAFLD_Diagnosis_AMC, predictor = data$Endo_conc)

png("ROC_NAFLD_C1.png", width = 600, height = 400)
# Plot the ROC curve
plot(roc_obj, col = "#428bca", lwd = 2,
     main = "ROC Curve: RNA Concentration for NAFLDdiagnostic Prediction")
# Add a diagonal line representing random classification
abline(a = 0, b = 1, lty = 2, col = "gray")

# Calculate and print the AUC value
auc_value <- auc(roc_obj)
print(paste("The AUC is:", round(auc_value, 3)))
dev.off()
```

